---
title: "R Users Group: Session 2 (dplyr)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Links to sections: 

- [Introduction](#introduction)
- [Cheatsheet](#cheatsheet)
- [Exercises](#exercises)




***********
# <a id="introduction"></a>INTRODUCTION: 
***********

## What is dplyr?

In this session we will be focusing on **data manipulation** with the `dplyr` package. Dplyr is one of the packages within the `tidyverse` group, which contains a number of core R packages including `dplyr`, `ggplot2`, `tidyr`, `purrr` amongst others. The tidyverse is a group of packages which are designed to simplify data science and share a common design philosophy, 'grammar' and data structures (see here for more information about the tidyverse: https://www.tidyverse.org/).

`dplyr` provides an alternative, and potentially more efficient/intuitive way to do data manipulation without using baseR commands.The `dplyr` package provides a 'grammar' of data manipulation and comes with 5 core functions which can be used to solve most data manupulation challenges:  

For more information on dplyr: https://dplyr.tidyverse.org/

- `filter()`    : filters the observations (rows) of a dataframe; like 'filter' in Excel  
- `arrange()`   : sets the order of observations (rows) according to a particular variable (column); like 'sort' in Excel
- `select()`    : used to select variables (columns) of a dataframe, useful for refining a dataframe to only the variables of interest
- `mutate()`    : creates new variables as a function of existing variables
- `summarise()` : provides summary statistics for a given variables (often used with `group_by` to produce grouped summary statistics)

**Support/resources**

An excellent introduction to data manipulation and transformation in R, using the `tidyverse` and `dplyr`, is given in Chapter 5 of the R for [Data Science textbook](https://r4ds.had.co.nz/transform.html). In today's session we will be working through the exercises from this book.

|  
|  
|  

************
# <a id="cheatsheet"></a>CHEAT SHEET
************

## Important points

1. Throughout these excercises, you can use the 'pipe' operator %>% which allows you to write clear code which can be read from left to right. The pipe takes the result of whatever is on the left of it, and 'pipes' it into the next function.


                x %>% f(y) becomes f(x, y)
                e.g. 
                iris %>% select(Petal.Length) becomes select(iris,Petal.Length)  

2. Note that dplyr uses a new data type: the tibble, which is a modern reimagining of the dataframe used within tidyverse packages. For more info see https://tibble.tidyverse.org/

## Example syntax:
- The examples below use the `starwars` dataset.

### `filter()` syntax
- Reminder: You may need to read about Comparisons and Logical Operators to understand how conditions can be applied with the filter() function. 
- See also `between()`
- Sections 5.2.1 and 5.2.2 of the [R for Data Science book](https://r4ds.had.co.nz/transform.html) provide more details

```{r eval = FALSE}
# filter by two levels:
starwars %>% filter(skin_color == "light", eye_color == "brown")

# Note that this is the same as: 
filter(starwars, skin_color == "light", eye_color == "brown")

# And similar to r base code:
starwars[starwars$skin_color == "light" & starwars$eye_color == "brown", ]

# When we add logical operators e.g. eye colour == blue OR brown, we write like so:
starwars %>% filter(eye_color == "blue" | eye_color == "brown")

 # OR using %in% :
starwars %>% filter(eye_colour %in% c("blue", "brown"))

```

### `arrange()` syntax
    
```{r eval = FALSE}
starwars %>% arrange(height, mass)

# to arrange in descending order :
starwars %>% arrange(desc(height))

```

### `select()` syntax
- Helper functions: `starts_with()`, `ends_with()`, `matches()` and `contains()`

```{r eval = FALSE}
# select the hair_color, skin_color and eye_color variables
starwars %>% select(hair_color, skin_color, eye_color)

# select all variables between hair_color and eye_color
starwars %>% select(hair_color:eye_color)

# select all BUT this variable
starwars %>% select(-hair_color)

```

### `mutate()` syntax
```{r eval = FALSE}
# create a new variable called height_m converting height from cm to m
starwars %>% mutate(height_m = height / 100) #note: you will need to assign this to a new dataframe to view the resulting variable

# similar to baseR code: 
starwars$height_metres2 <- starwars$height /100

```

### `summarise()` syntax
- You will often want to use in conjunction with `group_by()`
```{r eval = FALSE}
starwars %>% summarise(height = mean(height, na.rm = TRUE))

#with group_by
starwars %>% group_by(species) %>% summarise(height = mean(height, na.rm = TRUE))

```
|  
|  
|  
********
# <a id="exercises"></a>EXERCISES 
********

The solutions to these exercises can be found here: https://jrnold.github.io/r4ds-exercise-solutions/transform.html

### Set up

1. Load the `dplyr` package into the R workspace.  
*Note: dplyr is contained within the tidyverse package, you can therefore choose whether to load the full tidyverse or only dplyr for this session*

2. Load your data (we will use the publicly available `'flights'` dataset)  
- *Note: the `flights` dataset is part of the `nycflights13` package. You will need to install this package and load the library before you can access the data*
- You may wish to explore the dataset to remind yourselves of its contents using some **baseR** commands e.g. `head()`, `View()`, `str()`  

```{r eval=FALSE}

#install.packages("tidyverse")
#install.packages("dplyr")

library(tidyverse)
library(dplyr)


#Load the flights data

# install.packages("nycflights13")
library(nycflights13)
head(flights)

```


### `filter()`

1.  Find all flights that:

    - Had an arrival delay of two or more hours
    - Flew to Houston (`IAH` or `HOU`)
    - Departed in summer (July, August, and September)
    - Arrived more than two hours late, but didn't leave late
    - Departed between midnight and 6am (inclusive)

2.  How many flights have a missing `dep_time`? What other variables are 
    missing? What might these rows represent?

 
### `arrange()`

1.  How could you use `arrange()` to sort all missing values to the start?
    (Hint: use `is.na()`).
    
1.  Sort `flights` to find the most delayed flights.

1.  Sort `flights` to find the fastest (highest speed) flights.

1.  Which flights travelled the farthest? Which travelled the shortest?



### `select()`

1.  Brainstorm as many ways as possible to select `dep_time`, `dep_delay`,
    `arr_time`, and `arr_delay` from `flights`. (Hint: think about using the select helper functions)
    
1.  What happens if you include the name of a variable multiple times in
    a `select()` call?
  
1.  What does the `one_of()` function do? Why might it be helpful in conjunction
    with this vector?
    
    ```{r}
    vars <- c("year", "month", "day", "dep_delay", "arr_delay")
    ```
    
1.  Does the result of running the following code surprise you?  How do the
    select helpers deal with case by default? How can you change that default?

    ```{r, eval = FALSE}
    select(flights, contains("TIME"))
    ```



### `mutate()`


```{r, eval = FALSE, echo = FALSE}
flights <- flights %>% mutate(
  dep_time = hour * 60 + minute,
  arr_time = (arr_time %/% 100) * 60 + (arr_time %% 100),
  airtime2 = arr_time - dep_time,
  dep_sched = dep_time + dep_delay
)

ggplot(flights, aes(dep_sched)) + geom_histogram(binwidth = 60)
ggplot(flights, aes(dep_sched %% 60)) + geom_histogram(binwidth = 1)
ggplot(flights, aes(air_time - airtime2)) + geom_histogram()
```

1.  Currently `dep_time` and `sched_dep_time` are convenient to look at, but
    hard to compute with because they're not really continuous numbers. 
    Convert them to a more convenient representation of number of minutes
    since midnight.
    
```{r}
# To speed things up in this session, below is a piece of code which writes a function for converting time in 24 hour HH:MM format to minutes past midnight. You can now use this function within a 'mutate' argument. (This answer was taken from the unofficial answers sheet but can be achieved using packages made specifically for this task e.g. 'libridate')

time2mins <- function(x) {
  (x %/% 100 * 60 + x %% 100) %% 1440
}
```
    
1.  Compare `air_time` with `arr_time - dep_time`. What do you expect to see?
    What do you see? What do you need to do to fix it?
    
1.  Compare `dep_time`, `sched_dep_time`, and `dep_delay`. How would you
    expect those three numbers to be related?

1.  Find the 10 most delayed flights using a ranking function. How do you want 
    to handle ties? Carefully read the documentation for `min_rank()`.


### `summarise()`
1.  Brainstorm at least 5 different ways to assess the typical delay 
    characteristics of a group of flights. Consider the following scenarios:
    
    * A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of 
      the time.
      
    * A flight is always 10 minutes late.

    * A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of 
      the time.
      
    * 99% of the time a flight is on time. 1% of the time it's 2 hours late.
    
    Which is more important: arrival delay or departure delay?

1.  Come up with another approach that will give you the same output as 
    `not_cancelled %>% count(dest)` and 
    `not_cancelled %>% count(tailnum, wt = distance)` (without using 
    `count()`).

1.  Our definition of cancelled flights (`is.na(dep_delay) | is.na(arr_delay)`
    ) is slightly suboptimal. Why? Which is the most important column?

1.  Look at the number of cancelled flights per day. Is there a pattern?
    Is the proportion of cancelled flights related to the average delay?

1.  Which carrier has the worst delays? Challenge: can you disentangle the
    effects of bad airports vs. bad carriers? Why/why not? (Hint: think about
    `flights %>% group_by(carrier, dest) %>% summarise(n())`)

1.  What does the `sort` argument to `count()` do. When might you use it?
