---
title: "Session 1 (Introduction) with solutions"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

In this "Introduction to R" session we will split in to small groups to work through some basic exercises using the measles dataset\*. The data is based on WHO data for measles and rubella and has been adapted to contain information for measles only. The exercises are designed to direct some basic, exploratory data analysis and practice foundational skills in using R. When working through the exercises please do discuss within your group and support people with different experience levels. Once each group has completed the task we will have a short feedback session to discuss the different approaches used for each exercise. 

\*Official data based on monthly data reported to WHO (Geneva) as of May 2020, see here for original data: https://www.who.int/immunization/monitoring_surveillance/burden/vpd/surveillance_type/active/measles_monthlydata/en/

### Exercises: 
**1. Import your data**  

- Import the measles dataset into the R environment, assigning it to an appropriate variable  
*Hint: Set your working directory to the folder where the data is stored as your first step*

```{r}

#Set working directory:
setwd("~/R_users_group")

#Read in dataset and assign to an appropriate variable "measles_data"
measles_data <- read.csv("measlesreportedcasesbycountry.csv")

```

**2. Explore the dataset:**  

- What information is contained within the dataset?  
- What variables are there within the dataset and what type of data is stored within each column?  
- What are the dimensions of the dataset, what do these numbers correspond to?  
- Provide some summary descriptive statistics for the dataset  
- How many listed countries are in each region? (Tabulate)


```{r eval = FALSE}

#What information is contained within the dataset?
View(measles_data) #used to view the dataset in your R environment
head(measles_data)

#What variables are there within the dataset and what type of data is stored within each column
colnames(measles_data)
str(measles_data)
lapply(measles_data, class)

#What are the dimensions of the dataset and what do these numbers correspond to?
dim(measles_data) #rows (countries) and columns (variables)

#Provide some summary stats for the dataset
summary(measles_data)

#How many listed countries are in each region?
table(measles_data$Region)

```


**3. Further investigations**

- What's the mean and median number of confirmed measles cases in 2019? (remember to add the argument "na.rm" to remove NAs!)
- What's the median number of confirmed measles cases in 2019 for each Region? (This can be excecuted in a single line of code)
- Which country recorded the highest number of measles cases in 2019 and how many cases were recorded?  
- How many countries had 0 measles cases in 2019?  
- How many cases has Belgium confirmed so far in 2020, store your answer in the variable "Belgium_cases_2019"?  
- Add a column to the dataset which contains the proportion of suspected cases which have been confirmed for 2020  

```{r eval=FALSE}
#What's the mean and median number of confirmed measles cases in 2019? (remember to add the argument "na.rm" to remove NAs!) 
mean(measles_data$confirmed_cases_2019, na.rm = TRUE)
median(measles_data$confirmed_cases_2019, na.rm = TRUE)

#What's the median number of confirmed measles cases in 2019 for each Region? (This can be excecuted in a single line of code)
aggregate(confirmed_cases_2019 ~  Region, measles_data, median)

#Which country had the highest number of cases in 2019?
max_cases_2019 <- which.max(measles_data$confirmed_cases_2019) #find the index
measles_data[max_cases_2019, 2] #which country?
measles_data[max_cases_2019, "confirmed_cases_2019"] #how many cases

#How many countries recorded no measles cases in 2019
No_cases_2019 <- length(which(measles_data$confirmed_cases_2019 == 0))

#How many cases has Belgium confirmed so far in 2020?
Belgium_cases_2020 <- measles_data[measles_data$member_state == "Belgium", "suspected_cases_2020"]

#How many countries recorded have recorded fewer cases than Belgium so far in 2020?
length(which(measles_data$suspected_cases_2020<Belgium_cases_2020))

#Add a new column with the proportion of suspected cases that have been confirmed for 2020
measles_data$proportion_confirmed_2020 <- (measles_data$confirmed_cases_2020)/(measles_data$suspected_cases_2020)
```


**4. Investigate data for the Americas**  

- Create a new dataframe containing only information for the Americas (AMRO) region  
- Remove the data for 2019 and remove all rows containing NAs  
- Recode the data in the column for suspected cases as a binary variable, with values above the median coded as "high" and values below the median coded as "low".  
- How many countries are "high" and "low" and what proportion? (Tabulate the data)
*You may wish to copy the column of data first and work on the copied data*

```{r eval=FALSE}
#Create a new dataframe containing only the Americas region
americas <- measles_data[measles_data$Region=="AMRO", ]
# or alternatively
americas <- subset(measles_data, Region == "AMRO")

#Remove the data for 2019
americas <- americas[,-c(6:7)]
#or alternatively
americas <- subset(americas, select = -c(confirmed_cases_2019, incidence_rate_per_1mil_2019))

#Remove NAs
americas <- na.omit(americas)

#Recode suspected cases as binary variable
americas_median <- median(americas$suspected_cases_2020)
americas$suspected_cases_binary <- americas$suspected_cases_2020
americas$suspected_cases_binary <- ifelse(americas$suspected_cases_binary>=americas_median, "High", "Low")

#What type of data is this, can you plot it as a barplot?
plot(as.factor(americas$suspected_cases_binary)) #Class of column is character data, so must be converted to a factor for plotting

#How many countries are "high" and "low" and what proportion? (Tabulate the data)
americas_table <- table(americas$suspected_cases_binary)
prop.table(americas_table)

```


**5. Saving your work **  

- Save your script 
- Save your 'workspace' (all of the R objects you have created in this session)

```{r eval=FALSE}
# Save your script 
# Just use File > Save as

# Save your 'workspace' (all of the R objects you have created in this session)
save.image("session1.Rdata")

# When you close down R, it will ask you: "Save workspace image to C:/Program Files.../RData?" Click NO
# otherwise you will not have a nice blank session to work with when you begin a new session. 

```
